---
title: System-level analysis of Alzheimer’s disease prioritizes candidate genes
  for neurodegeneration
author: "Jeffrey L. Brabec, Montana Kay Lara, Anna L. Tyler, J. Matthew Mahoney"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
bibliography
---

This workflow accompanies the paper "System-level analysis of Alzheimer's disease prioritizes candidate genes for neurodegeneration". It contains all of the data and code to replicate the results and figures in the manuscript with a few key exceptions. The results presented in the manuscript were partially generated using data from the Alzheimer's Disease Neuroimaging Initiative [(ADNI)](http://adni.loni.usc.edu). We cannot provide SNP-level summary statistics, nor can we provide the phenotypic information used to generate figures 2A and 2B. However, the rest of the R analysis pipeline is included from the gene-level summary statistics. ADNI summary statistics were calculated by us using PLINK 1.9 (Cite Chang et al. 2015) while the meta-GWAS summary statistics were obtained from the website of [Jansen et al. 2019](https://ctg.cncr.nl/documents/p1651/AD_sumstats_Jansenetal_2019sept.txt.gz), the lab that generated them.

The purpose of this project was to develop a novel method integrating GWAS statistics and functional scores, derived from network-based functional prediction (NBFP), to prioritize novel Alzheimer's Disease (AD) genes. The majority of AD GWAS studies conducted to date have utilized case-control data to identify single nucleotide polymorphisms (SNPs) that segregate with disease. These analyses have identified several genes that drive most of the heritability of AD, such as *APOE*, *APP*, and *PSEN1/2*, however, they do not have the biological resolution to identify variants that could be driving tissue or cell level processes that lead to AD phenotypes such as neurodegeneration, reactive astrocytosis, and the build-up of amyloid-plaques and tau-tangles.

We therefore decided to utilize brain-region volume phenotypes available through the ADNI to run a GWAS to identify genes which were associated with changes in hippocampal and amygdalar volume. Our ultimate goal was to identify novel gene candidates for AD-specific neurodegeneration.

Our methodology prioritizes genes based on their functional significance, derived from our NBFP pipeline and based on tissue-relevance (e.g. hippocampus or amygdala). This method has been adapted from the work flow developed in Tyler et al. 2019, and based on the methodology described in Guan et al. 2010 in which SVMs were used to predict functional associations with mammalian phenotype (MP) ontology terms. We use SVMs to determine which genes are functionally associated with AD in specific hippocampal or amygdalar contexts.

[Expand on important details and trim other bits]

## I. Setup

This analysis requires a range of R packages, some of which are available from [CRAN](https://cran.r-project.org) and some of which are available from [Bioconductor](http://bioconductor.org). We've included a comprehensive list of the packages below how to install them. **Please note that this workflow was last updated in R 4.0.3 and has not been tested in later versions of the software.**

### Package installation

```{r install, include = FALSE}
cran_pkgs <- c("tidyverse",
               "here",
               "conflicted",
               "gprofiler2",
               "R.utils",
               "doParallel",
               "DescTools",
               "knitr",
               "igraph",
               "RColorBrewer",
               "plotly",
               "ggrepel",
               "janitor",
               "tidytext",
               "ggpubr",
               "ggpmisc",
               "viridis",
               "scales")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(pkgs = "biomaRt")
```

### A. Package Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
all.packages <- c("tidyverse", 
                  "here", 
                  "conflicted", 
                  "gprofiler2", 
                  "R.utils", 
                  "doParallel", 
                  "DescTools", 
                  "knitr",
                  "igraph",
                  "RColorBrewer", 
                  "biomaRt",
                  "plotly",
                  "ggrepel",
                  "janitor",
                  "tidytext",
                  "ggpubr",
                  "ggpmisc",
                  "viridis",
                  "scales")
for(i in 1:length(all.packages)){library(all.packages[i], character.only = TRUE)} 
conflict_prefer("filter", "dplyr")
```

### B. Directory Setup

```{r}
project.dir <- here()
code.dir <- here("code")
data.dir <- here("data")
results.dir <- here("results")

#MetaGWAS Results Directories
results.mwas.amygdala.net.dx.dir <- here("mwas", "mwas_amygdala_net_dx")
results.mwas.hippocampus.net.dx.dir <- here("mwas", "mwas_hippocampus_net_dx")
```

### C. Source the Code Used in the Project

This chunk will read in all of the custom functions written to format the networks and run the machine learner.

```{r source_code}
all.fun.dir <- list.files(code.dir, full.names = TRUE)
for(i in 1:length(all.fun.dir)){
  all.fun <- list.files(all.fun.dir[i], full.names = TRUE)
  for(j in 1:length(all.fun)){source(all.fun[j])}
}
```

### D. Download Tissue Networks

Download HumanBase functional tissue network. The specified networks used in this analysis were downloaded on July 13, 2020. The function `download.tissue.network()` will download the specified network (top edges) and place it in the specified directory. To view all available networks and naming paradigm visit [HumanBase](https://hb.flatironinstitute.org/download).

```{r download_tiss_nets}
hum_hippocampus.net <- here("data", "hippocampus_top.Rdata")
if(!file.exists(hum_hippocampus.net)){
  download.tissue.net(tissue = "hippocampus", organism = "human", project.dir = data.dir)
}

hum_amygdala.net <- here("data", "amygdala_top.Rdata")
if(!file.exists(hum_amygdala.net)){
  download.tissue.net(tissue = "amygdala", organism = "human", project.dir = data.dir)
}

```

### E. Set SVM Hyper-Parameters

We chose a value of 100 SVMs based on the work done in Guan et al. (2010), however a higher or lower number can be specified if desired. We've also allowed the user to specify the number of cores to use for running the models as its computationally intensive and can severely slow computers with less processing power.

```{r}
#set paramaters
n.SVM <- 100 #number of SVMs to run
n.cores <- detectCores()-1 #The number of cores to use for parallel processing

```

## II. Genes

### A. Load MAGMA Results

SNP-level summary statistics were retrieved from the ADNI and Jansen et al. 2019 as stated above. We mapped SNP-level statistics to gene-level using Multi-marker Analysis of GenoMic Annotation [(MAGMA)](https://ctg.cncr.nl/software/magma). We used a 10kb annotation window around each gene and used the NCBI hg37 reference genome. LD estimations were done using the 1000 Genome's project CEU population. The MAGMA output has columns for gene, chromosome, start and stop positions, number of SNPs, N number of samples, Z statistic, and p-value. The gene-level summary statistics are provided in the GitHub repository along with this workflow, and will be read into the work-space with the following chunk.

```{r}
#Hippocampal Volume
hv_adni_genes <- read_table(here("data", "hipp_10k_genes.genes.out"), 
                            col_names = TRUE) %>% 
  mutate(across(starts_with("GENE"), as.character)) %>% 
  clean_names()

#Amygdala Volume
av_adni_genes <- read_table(here("data", "amyg_10k_genes.genes.out"),
                            col_names = TRUE) %>% 
  mutate(across(starts_with("GENE"), as.character)) %>% 
  clean_names()

#MetaGWAS 
mwas_genes <- read_table(here("data", "sum_stats_genes.genes.out"),
                         col_names = TRUE) %>% 
  mutate(across(starts_with("GENE"), as.character)) %>% 
  clean_names()
```

### B. Gene Entrez ID Conversion

The networks that we obtain from [HumanBase](https://hb.flatironinstitute.org) name genes using Entrez IDs, however for visualization and follow-up we'll need the actual gene symbols. To obtain the gene symbols we use the `gprofiler2` function `gconvert()` to convert the gene Entrez IDs from the MAGMA output to gene symbols. Then, the original data is joined together with the converted gene names to get the final dataset to be used for down-stream processing.

```{r}
#Amygdala Volume
av_adni_gconvert <- gconvert(av_adni_genes$gene,
                             organism = "hsapiens",
                             target = "ENTREZGENE",
                             numeric_ns = "ENTREZGENE_ACC") %>% 
  filter(!duplicated(name), !duplicated(input)) %>% 
  as_tibble() %>% 
  dplyr::select(gene = input, name) %>% 
  mutate(across(gene, as.character))

av_adni_genes %<>% inner_join(av_adni_gconvert, by = "gene")

#Hippocampus Volume
hv_adni_gconvert <- gconvert(hv_adni_genes$gene,
                             organism = "hsapiens",
                             target = "ENTREZGENE",
                             numeric_ns = "ENTREZGENE_ACC") %>% 
  filter(!duplicated(name), !duplicated(input)) %>% 
  as_tibble() %>% 
  dplyr::select(gene = input, name) %>% 
  mutate(across(gene, as.character))

hv_adni_genes %<>% inner_join(hv_adni_gconvert, by = "gene")

#MetaGWAS
mwas_gconvert <- gconvert(mwas_genes$gene,
                          organism = "hsapiens",
                          target = "ENTREZGENE",
                          numeric_ns = "ENTREZGENE_ACC") %>% 
  filter(!duplicated(name), !duplicated(input)) %>% 
  as_tibble() %>% 
  dplyr::select(gene = input, name) %>%  
  mutate(across(gene, as.character))

mwas_genes %<>% inner_join(mwas_gconvert, by = "gene")
```

## III. Significant Genes (Positional Candidates)

Find significant genes by filtering the gene list for genes with a nominal p-value below 0.01. These genes will be used to train the SVM. (Note: only gene Entrez IDs are needed for training) These genes represent positional candidates.

```{r sig_genes}
mwas_sig_genes <- mwas_genes %>% 
  filter(p < 0.01)
```

## IV. SVM

We ran the SVM analysis on the meta-GWAS genes using the amygdala network and then the hippocampus network. The SVMs were trained to distinguish genes in the significant gene list from randomly chosen genes from outside the list. We use the function generate.triage.models() to accomplish this, which uses the `svm` function in the package `e1071` [@e1071]. The features used by the SVM are the edge weights between all genes and the genes in the meta-GWAS significant gene set. For each of the 100 SVM models, we used the meta-GWAS significant gene set genes as the true positives, while the true negatives were selected at random from outside of that gene set. Each true negative set was selected to have the same number of genes as were in the true positive set.

### A. Train SVM using the significant genes and the specified functional network

**Be aware that the final scored tables are stored in the sub-directories of the `mwas`** **directory and can be read in under the Section V. B code chunk below. The interceding chunks which calculate the ROC curves, score the models, and merge the SVM results to the GWAS data have `if`** **statements which will prevent premature running of the chunks as they are memory intensive.**

```{r train_SVM}
if(FALSE){
##read in amygdala network
amygdala_net <- read_rds(here("data", "amygdala_top.RData"))


#1. MWAS Diagnosis | Amygdala Network
generate.triage.models(path = results.mwas.amygdala.net.dx.dir, 
                       project.name = "results_mwas_amygdala_net_dx", 
                       trait.genes = mwas_sig_genes$GENE,
                       fgn = amygdala_net, 
                       n.trials = n.SVM, 
                       cluster.modules = FALSE, 
                       use.SVD = FALSE, 
                       verbose = TRUE, 
                       n.cores = n.cores)
##remove amygdala network once svms have run to free up RAM
rm(amygdala_net)

##read in hippocampus network
hippocampus_net <- here(data.dir, "hippocampus_top.RData")

#2. MWAS Diagnosis | Hippocampus Network
generate.triage.models(path = results.mwas.hippocampus.net.dx.dir, 
                       project.name = "results_mwas_hippocampus_net_dx", 
                       trait.genes = mwas_sig_genes$GENE,
                       fgn = hippocampus_net, 
                       n.trials = n.SVM, 
                       cluster.modules = FALSE, 
                       use.SVD = FALSE, 
                       verbose = TRUE, 
                       n.cores = n.cores)

##remove hippocampus network once svms have run to free up RAM
rm(hippocampus_net)
}
```

### B. ROC curves

Following training of the SVM models, we evaluated them by generating ROC curves to test how well the models separate genes annotated to the network from genes in the rest of the genome. To do this, use the svm.ROC function, which plots the false positive rate against the true positive rate of the SVM model across the full range of prediction values. An area under the ROC curve (AUC) indicates how well the model does overall (AUC 0.5 means model performs at random chance, AUC 1 means model has perfect classification)

```{r ROC_curve}
if(FALSE){
svm.ROC(results.mwas.amygdala.net.dx.dir, n.cores = n.cores, plot.results = TRUE)
svm.ROC(results.mwas.hippocampus.net.dx.dir, n.cores = n.cores, plot.results = TRUE)
}
```

### C. Apply the trained models to genes

After training the SVM models, we use them to classify each gene, where each gene receives a score from each SVM indicating whether the model identifies it as part of the [network] gene list (a positive score) or not part of the gene list (negative score). The score.candidates function writes out all of the values to two csv tables (Candidate.Gene.SVM.Scores.csv and Candidate.Gene.FP.Rates.csv). The function also plots boxplots of the distributions of the SVM scores and the false positive (FP) rates for each gene in two jpgs (Candidate.Gene.SVM.Scores.jpg, and Candidate.Gene.FP.Rates.jpg). Plots are ordered by mean value. The genes that are on the left are classified confidently by the SVM as being related to the [network] gene list. The distributions of their SVM scores do not cross 0. The genes on the right hand side are classified confidently by the SVM as not being related to the [network] gene list. The genes in between are less confidently classified across the SVM trials.

```{r score_candidates}
if(FALSE){
score.candidates(results.mwas.amygdala.net.dx.dir, candidate.genes = mwas_genes$gene)
score.candidates(results.mwas.hippocampus.net.dx.dir, candidate.genes = mwas_genes$gene)
}
```

## V. Integrate Gene Information with SVM Output

We integrated the SVM scores and FP Rates with the original gene information (including positional information). The function `merge.svm.gene.info.fp()` takes the original gene table (output from MAGMA plus the converted gene names) and adds the mean SVM score and mean FP rate to the genes. It writes out a table for each module with the information for each gene, including the original information from the gene table (entrez ID and gene name), SVM score, and FP rate (Candidate.Gene.Results.rds).

### A. Merge SVM results with GWAS info

```{r merge_info}
if(FALSE){
#MWAS Diagnosis | Amygdala Network
mwas_amygdala_net_fin_tab <- merge.gene.svm.fp(results.mwas.amygdala.net.dx.dir, mwas_genes, gene.column.name = "gene")

#MWAS Diagnosis | Hippocampus Network
mwas_hippocampus_net_fin_tab <- merge.gene.svm.fp(results.mwas.hippocampus.net.dx.dir, mwas_genes, gene.column.name = "gene")
}
```

### B. Read in already merged tables

```{r}
mwas_amygdala_net_fin_tab <- 
  read_rds(here("results/Amygdala.Candidate.Gene.Results.rds"))

mwas_hippocampus_net_fin_tab <- 
  read_rds(here("results/Hippocampus.Candidate.Gene.Results.rds"))
```

## VI. Gene Scores and Plotting

Each gene is scored based on the integration of SVM false positive (FP) rates, GWAS p-values, and potentially other metrics given the dataset (e.g., expression data). Given two variables (e.g., -log p-value and -log SVM FP rate), the score is based on the number of genes in the area covered by its x,y position. An individual gene with a lot of genes covered in their area will be ranked higher than a gene with fewer. Given only positional information and SVM output, genes can be ranked on their FP rates and mapped by their position (and significant GWAS genes can also be designated from the rest).

## A. Plotting Integrated Ranking Within GWAS Method (Fig4)

```{r integrated_rankings}

#Extract the gene names and p-values from the Amygdalar Volume GWAS
amyg_adni_scores <- av_adni_genes %>% 
  dplyr::select(name, av_p = p)

#Extract the gene names and the Amygdalar functional scores for the meta-GWAS
mwas_amyg_scores <- mwas_amygdala_net_fin_tab %>% 
  dplyr::select(name, mwas_amyg_fp = mean_fp_rate)

#Join the two amygdalar datasets and calculate the combined score.
amygdala_scores <- amyg_adni_scores %>% 
  inner_join(mwas_amyg_scores, by = "name") %>% 
  mutate(
    av_log_p = -log10(av_p),
    mwas_log_amyg_fp = -log10(mwas_amyg_fp)
  ) %>% 
  rowwise() %>%
  mutate(comb_score_amyg = sum(av_log_p > .$av_log_p & mwas_log_amyg_fp > .$mwas_log_amyg_fp)) %>%
  ungroup() %>%
  mutate(comb_score_amyg = comb_score_amyg/n())

#Extract gene names and Hippocampal Volume GWAS p-values
hipp_adni_scores <- hv_adni_genes %>% 
  dplyr::select(name, hv_p = p)

#Extract gene names and the Hippocampal functional scores for the meta-GWAS
mwas_hipp_scores <- mwas_hippocampus_net_fin_tab %>% 
  dplyr::select(name, mwas_hipp_fp = mean_fp_rate)

#Join the two hippocampal datasets and calculate the combined score.
hippocampus_scores <- hipp_adni_scores %>% 
  inner_join(mwas_hipp_scores, by = "name") %>% 
  mutate(
    hv_log_p = -log10(hv_p),
    mwas_log_hipp_fp = -log10(mwas_hipp_fp)
    ) %>%
  rowwise() %>%
  mutate(comb_score_hipp = sum(hv_log_p > .$hv_log_p & mwas_log_hipp_fp > .$mwas_log_hipp_fp)) %>%
  ungroup() %>%
  mutate(comb_score_hipp = comb_score_hipp/n())


#1. Generate Amygdala tissue combined plot using custom plotting function
amyg_scores_comb_plot <- gene_rank_plotr(score_tab = amygdala_scores,
                gene_col = "name",
                log_pval_col = "av_log_p",
                log_score_comp_col = "mwas_log_amyg_fp",
                set_ceiling = FALSE) +
  labs(x = "Amygdala Volume Positional Score",
       y = "Functional Score",
       title = "Amygdala Network",
       color = "Combined\nScore") +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 36),
    legend.text = element_text(size = 25),
    axis.title.y = element_text(size = 36),
    axis.title.x = element_text(size = 36),
    axis.text.y = element_text(size = 25),
    axis.text.x = element_text(size = 25),
    title = element_text(size = 42),
    panel.grid = element_line(color = "#000000"),
    panel.grid.minor = element_blank()
    )

#2. Generate Hippocampal tissue combined plot using custom plotting function
hipp_scores_comb_plot <- gene_rank_plotr(score_tab = hippocampus_scores,
                gene_col = "name",
                log_pval_col = "hv_log_p",
                log_score_comp_col = "mwas_log_hipp_fp",
                set_ceiling = FALSE) +
  labs(x = "Hippocampus Volume Positional Score",
       y = "Functional Score",
       title = "Hippocampus Network",
       color = "Combined\nScore") +
  expand_limits(x = 16) +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 36),
    legend.text = element_text(size = 25),
    axis.title.y = element_text(size = 36),
    axis.title.x = element_text(size = 36),
    axis.text.y = element_text(size = 25),
    axis.text.x = element_text(size = 25),
    title = element_text(size = 42),
    panel.grid = element_line(color = "#000000"),
    panel.grid.minor = element_blank()
    )

## Interactive plotly plots:
gene_rank_plotlyr(score_tab = amygdala_scores,
                  gene_col = "name",
                  log_p_col = "av_log_p",
                  log_comp_col = "mwas_log_amyg_fp",
                  title = "Amygdala Interactive Ranking Plot")

gene_rank_plotlyr(score_tab = hippocampus_scores,
                  gene_col = "name",
                  log_p_col = "hv_log_p",
                  log_comp_col = "mwas_log_hipp_fp",
                  title = "Hippocampus Interactive Ranking Plot")
```

### B. Generating P-value Comparison Plots

#### Comparing ADNI Volume GWAS P-values (Fig 2C)

```{r}

#Join the amygdala and hippocampus score data.frames generated above and calculate a combined ranking for plot color and text labelling purposes.
amyg_hipp_scores <- amygdala_scores %>% 
  inner_join(hippocampus_scores, by = "name") %>% 
  select(name, av_log_p, hv_log_p, mwas_log_amyg_fp, mwas_log_hipp_fp, comb_score_amyg, comb_score_hipp) %>% 
  rowwise() %>%
  mutate(comb_score = sum(hv_log_p > .$hv_log_p & av_log_p > .$av_log_p)) %>%
  ungroup() %>%
  mutate(comb_score = comb_score/n())
  
#Generate the comparison plot! Colors were taken from the viridis package palette.
amyg_hipp_pval_comparison <- amyg_hipp_scores %>% 
  ggplot(aes(hv_log_p, av_log_p)) +
  geom_point(color = "grey", size = 5) +
  geom_point(data = amyg_hipp_scores %>% 
               filter(hv_log_p > -log10(0.01)), color = "#F0F921FF", size = 6) +
  geom_point(data = amyg_hipp_scores %>% 
               filter(av_log_p > -log10(0.01)), color = "#F48849FF", size = 6) +
  geom_point(data = amyg_hipp_scores %>% 
               filter(hv_log_p > -log10(0.01) & av_log_p > -log10(0.01)), color = "#5402A3FF", size = 6) +
  geom_text_repel(data = amyg_hipp_scores %>% 
                    filter(av_log_p < -log10(0.01)) %>% 
                    slice_max(hv_log_p, n = 5), 
                  aes(label = name, fontface = "italic"), 
                  size = 10,
                  nudge_x = 4, #for hippocampus
                  direction = "y",
                  xlim = c(5, Inf)) +
  geom_text_repel(data = amyg_hipp_scores %>% 
                    filter(hv_log_p < -log10(0.01)) %>% 
                    slice_max(av_log_p, n = 5), 
                  aes(label = name, fontface = "italic"), 
                  size = 10,
                  nudge_y = 4,
                  direction = "x",
                  ylim = c(5, Inf)) +
  geom_text_repel(data = amyg_hipp_scores %>% 
                    filter(hv_log_p > -log10(0.01) & av_log_p > -log10(0.01)) %>% 
                    slice_max(comb_score, n = 5), 
                  aes(label = name, fontface = "italic"), 
                  size = 10,
                  xlim = c(6.5, Inf)) +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", size = 3) +
  geom_vline(xintercept = -log10(0.01), linetype = "dashed", size = 3) +
  labs(
    x = "Hippocampus Volume GWAS -log10 P-Value",
    y = "Amygdala Volume GWAS -log10 P-Value"
  ) +
  expand_limits(x = 16) +
  theme_minimal() +
  theme(
    axis.title.y = element_text(size = 36),
    axis.title.x = element_text(size = 36),
    axis.text.y = element_text(size = 25),
    axis.text.x = element_text(size = 25),
    panel.grid.major = element_line(color = "#000000"),
    panel.grid.minor = element_blank()
  )
```

#### Comparing MWAS p-values to ADNI Volume GWAS p-values (Figs 2D and E)

```{r}
mwas_genes_to_bind <- mwas_genes %>% 
  dplyr::select(name, mwas_p = p)

av_adni_genes_to_bind <- av_adni_genes %>% 
  dplyr::select(name, av_p = p)

hv_adni_genes_to_bind <- hv_adni_genes %>% 
  dplyr::select(name, hv_p = p)

hv_mwas_comp_tab <- hv_adni_genes_to_bind %>% 
  inner_join(mwas_genes_to_bind, by = "name") %>% 
  mutate(
    hv_log_p = -log10(hv_p),
    mwas_log_p = -log10(mwas_p)
  )

av_mwas_comp_tab <- av_adni_genes_to_bind %>% 
  inner_join(mwas_genes_to_bind, by = "name") %>% 
  mutate(
    av_log_p = -log10(av_p),
    mwas_log_p = -log10(mwas_p)
  )

av_mwas_comp_tab_processed <- av_mwas_comp_tab %>%
  rowwise() %>%
  mutate(comb_score = sum(av_log_p > .$av_log_p & mwas_log_p > .$mwas_log_p)) %>%
  ungroup() %>%
  mutate(comb_score = comb_score/n()) %>% 
  mutate(p_ceiling = ifelse(mwas_log_p > 20, 20, mwas_log_p))

av_mwas_p_value_comp_plot <- ggplot(av_mwas_comp_tab_processed,
                                    aes(x = av_log_p, y = p_ceiling, label = name)) +
  geom_point(color = "grey", size = 5) +
  geom_hline(yintercept = 19, linetype = "dashed", color  = "grey50", size = 3) +
  geom_text(aes(x = 8, y = 18, label = "Genes with -log10 P > 20 transformed to 20 for readability"), size = 10, color = "grey50") +
  geom_point(data = subset(av_mwas_comp_tab_processed, av_log_p > -log10(0.01)), color = "#F48849FF", size = 6) +
  geom_text_repel(data = av_mwas_comp_tab_processed %>% 
                    filter(mwas_log_p < -log10(0.01)) %>% 
                    slice_max(av_log_p, n = 5), 
                  aes(label = name, fontface = "italic"), 
                  size = 10,
                  nudge_x = 2,
                  direction = "y",
                  ylim = c(2, Inf)) +
  geom_point(data = subset(av_mwas_comp_tab_processed, mwas_log_p > -log10(0.01)), color = "#21908CFF", size = 6) +
  geom_text_repel(data = av_mwas_comp_tab_processed %>%
                    filter(av_log_p < -log10(0.01)) %>% 
                    slice_max(mwas_log_p, n = 5), 
                  aes(label = name, fontface = "italic"),
                  size = 10,
                  nudge_y = 4,
                  direction = "y",
                  xlim = c(2, Inf)) +
  geom_point(data = av_mwas_comp_tab_processed %>% 
               filter(av_log_p > -log10(0.01) & mwas_log_p > -log10(0.01)), color = "#440154FF", size = 6) +
  geom_text_repel(data = av_mwas_comp_tab_processed %>% 
                    filter(av_log_p > -log10(0.01) & mwas_log_p > -log10(0.01)) %>% 
                    slice_max(comb_score, n = 5), 
                  aes(label = name, fontface = "italic"), 
                  size = 10) +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", size = 3) +
  geom_vline(xintercept = -log10(0.01), linetype = "dashed", size = 3) +
  labs(
    x = "ADNI Amygdala Volume GWAS -log10 P-value",
    y = "MetaGWAS -log10 P-value"
  ) +
  expand_limits(y = 21) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.y = element_text(size = 36),
    axis.title.x = element_text(size = 36),
    axis.text.y = element_text(size = 25),
    axis.text.x = element_text(size = 25),
    panel.grid.major = element_line(color = "#000000"),
    panel.grid.minor = element_blank()
  )


hv_mwas_comp_tab_processed <- hv_mwas_comp_tab %>%
  rowwise() %>%
  mutate(comb_score = sum(hv_log_p > .$hv_log_p & mwas_log_p > .$mwas_log_p)) %>%
  ungroup() %>%
  mutate(comb_score = comb_score/n()) %>% 
  mutate(p_ceiling = ifelse(mwas_log_p > 20, 20, mwas_log_p))

hv_mwas_p_value_comp_plot <- ggplot(hv_mwas_comp_tab_processed,
                                    aes(x = hv_log_p, y = p_ceiling, label = name)) +
  geom_point(color = "grey", size = 5) +
  geom_hline(yintercept = 19, linetype = "dashed", color  = "grey50", size = 3) +
  geom_text(aes(x = 8.5, y = 18, label = "Genes with -log10 P > 20 transformed to 20 for redability"), size = 10, color = "grey50") +
  geom_point(data = subset(hv_mwas_comp_tab_processed, hv_log_p > -log10(0.01)), color = "#F0F921FF", size = 6) +
  geom_text_repel(data = hv_mwas_comp_tab_processed %>% 
                    filter(mwas_log_p < -log10(0.01)) %>% 
                    slice_max(hv_log_p, n = 5), 
                  aes(label = name, fontface = "italic"), 
                  size = 10,
                  nudge_x = 3,
                  direction = "y",
                  ylim = c(2, Inf)) +
  geom_point(data = subset(hv_mwas_comp_tab_processed, mwas_log_p > -log10(0.01)), color = "#21908CFF", size = 6) +
  geom_text_repel(data = hv_mwas_comp_tab_processed %>%
                    filter(hv_log_p < -log10(0.01)) %>% 
                    slice_max(mwas_log_p, n = 5), 
                  aes(label = name, fontface = "italic"), 
                  size = 10,
                  nudge_y = 4,
                  direction = "y",
                  xlim = c(2, Inf)) +
  geom_point(data = hv_mwas_comp_tab_processed %>% 
               filter(hv_log_p > -log10(0.01) & mwas_log_p > -log10(0.01)), color = "#440154FF", size = 6) +
  geom_text_repel(data = hv_mwas_comp_tab_processed %>% 
                    filter(hv_log_p > -log10(0.01) & mwas_log_p > -log10(0.01)) %>% 
                    slice_max(comb_score, n = 5), 
                  aes(label = name, fontface = "italic"), 
                  size = 10) +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", size = 3) +
  geom_vline(xintercept = -log10(0.01), linetype = "dashed", size = 3) +
  labs(
    x = "ADNI Hippocampus Volume GWAS -log10 P-value",
    y = "MetaGWAS -log10 P-value"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.y = element_text(size = 36),
    axis.title.x = element_text(size = 36),
    axis.text.y = element_text(size = 25),
    axis.text.x = element_text(size = 25),
    panel.grid.major = element_line(color = "#000000"),
    panel.grid.minor = element_blank()
  )
```

#### Pie chart generation (Fig 5)

```{r}
amyg_comb_score_revigo <- read_delim(here("results/REVIGO_treemap_amyg_comb_converted.csv"), delim = "\t")

amyg_comb_score_revigo_for_plotting <- amyg_comb_score_revigo %>% 
  count(representative) %>% 
  rowwise() %>% 
  mutate(perc = n/sum(.$n)) %>% 
  mutate(plot_groups = ifelse(perc > .04, representative, "Other")) %>% 
  ungroup() %>% 
  count(plot_groups) %>% 
  rowwise() %>% 
  mutate(perc = n/sum(.$n)) %>% 
  ungroup() %>% 
  mutate(plot_groups = fct_reorder(plot_groups, perc*100))

amyg_comb_score_revigo_for_plotting %>% 
  ggplot(aes("", perc, fill = plot_groups)) + 
  geom_bar(width = 1, stat = 'identity') +
  geom_text(aes(x = 1.75, label = percent(perc)), 
            size = 7,
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start=0) +
  labs(title = "Amygdala Combined Score Functional Enrichment",
       fill = "GO Term and Proportion of Biological Process") +
  scale_fill_manual(values = cividis(n = 5)) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 20),
    panel.border = element_blank(),
    panel.grid=element_blank(),
    axis.ticks = element_blank(),
    plot.title=element_text(size=36, face="bold", hjust = 0.5),
    legend.title = element_text(size = 30, face = "bold"),
    legend.text = element_text(size = 25)
  )

amyg_pval_revigo <- read_delim(here("results/REVIGO_treemap_amyg_pval_converted.csv"), delim = "\t")

amyg_pval_revigo_for_plotting <- amyg_pval_revigo %>% 
  count(representative) %>% 
  rowwise() %>% 
  mutate(perc = n/sum(.$n)) %>% 
  mutate(plot_groups = ifelse(perc > .04, representative, "Other")) %>% 
  ungroup() %>% 
  count(plot_groups) %>% 
  rowwise() %>% 
  mutate(perc = n/sum(.$n)) %>% 
  ungroup() %>% 
  mutate(plot_groups = fct_reorder(plot_groups, perc*100))

amyg_pval_revigo_for_plotting %>% 
  ggplot(aes("", perc, fill = plot_groups)) + 
  geom_bar(width = 1, stat = 'identity') +
  geom_text(aes(x = 1.75, label = percent(perc)), 
            size = 7,
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start=0) +
  labs(title = str_wrap("Amygdala Volume GWAS P-value Functional Enrichment"),
       fill = "GO Term and Proportion of Biological Process") +
  scale_fill_manual(values = cividis(n = 7)) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 20),
    panel.border = element_blank(),
    panel.grid=element_blank(),
    axis.ticks = element_blank(),
    plot.title=element_text(size=36, face="bold", hjust = 0.5),
    legend.title = element_text(size = 30, face = "bold"),
    legend.text = element_text(size = 25)
  ) 

hipp_comb_score_revigo <- read_delim(here("results/REVIGO_treemap_hipp_comb_converted.csv"), delim = "\t")

hipp_comb_score_revigo_for_plotting <- hipp_comb_score_revigo %>% 
  count(representative) %>% 
  rowwise() %>% 
  mutate(perc = n/sum(.$n)) %>% 
  mutate(plot_groups = ifelse(perc > .04, representative, "Other")) %>% 
  ungroup() %>% 
  count(plot_groups) %>% 
  rowwise() %>% 
  mutate(perc = n/sum(.$n)) %>% 
  ungroup() %>% 
  mutate(plot_groups = fct_reorder(plot_groups, perc*100))

hipp_comb_score_revigo_for_plotting %>% 
  ggplot(aes("", perc, fill = plot_groups)) + 
  geom_bar(width = 1, stat = 'identity') +
  geom_text(aes(x = 1.75, label = percent(perc)), 
            size = 7,
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start=0) +
  labs(title = str_wrap("Hippocampus Combined Score Functional Enrichment"),
       fill = "GO Term and Proportion of Biological Process") +
  scale_fill_manual(values = plasma(n = 7)) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 20),
    panel.border = element_blank(),
    panel.grid=element_blank(),
    axis.ticks = element_blank(),
    plot.title=element_text(size=36, face="bold", hjust = 0.5),
    legend.title = element_text(size = 30, face = "bold"),
    legend.text = element_text(size = 25)
  ) 

hipp_pval_revigo <- read_delim(here("results/REVIGO_treemap_hipp_pval_converted.csv"), delim = "\t")

hipp_pval_revigo_for_plotting <- hipp_pval_revigo %>% 
  count(representative) %>% 
  rowwise() %>% 
  mutate(perc = n/sum(.$n)) %>% 
  mutate(plot_groups = ifelse(perc > .04, representative, "Other")) %>% 
  ungroup() %>% 
  count(plot_groups) %>% 
  rowwise() %>% 
  mutate(perc = n/sum(.$n)) %>% 
  ungroup() %>% 
  mutate(plot_groups = fct_reorder(plot_groups, perc*100))

hipp_pval_revigo_for_plotting %>% 
  ggplot(aes("", perc, fill = plot_groups)) + 
  geom_bar(width = 1, stat = 'identity') +
  geom_text(aes(x = 1.75, label = percent(perc)), 
            size = 7,
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start=0) +
  labs(title = str_wrap("Hippocampus Volume P-value Functional Enrichment"),
       fill = "GO Term and Proportion of Biological Process") +
  scale_fill_manual(values = plasma(n = 7)) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 20),
    panel.border = element_blank(),
    panel.grid=element_blank(),
    axis.ticks = element_blank(),
    plot.title=element_text(size=36, face="bold", hjust = 0.5),
    legend.title = element_text(size = 30, face = "bold"),
    legend.text = element_text(size = 25)
  ) 
```
